Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Community Detection and Parameter Estimation

### 2021-03-16 summary.Rmd
proof for theorem 4

---
title: "Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Community Detection and Parameter Estimation"
author: John Koo, Minh Tang, Michael Trosset
output:
  pdf_document:
    citation_package: natbib
    number_sections: yes
# output: html_document
fontsize: 12pt
# geometry: "left=1cm,right=1cm,top=1cm,bottom=1.5cm"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square}
- \usepackage{verbatim}
bibliography: misc.bib
abstract: |
  In this paper, we connect two random graph models, the Popularity 
  Adjusted Block Model (PABM) and the Generalized Random Dot Product Graph 
  (GRDPG) and use properties established in this connection to aid in 
  community detection and parameter estimation. In particular, we note that the
  PABM can be represented as latent positions such that points within the same 
  community lie on a subspace, and the subspaces that represent each 
  community are orthogonal to one another. Using this property as well as the 
  asymptotic properties of Adjacency Spectral Embedding (ASE) of the GRDPG, we 
  are able to establish theoretical asymptotic results of our community 
  detection and parameter estimation methods for the PABM. 
---

@@

## The popularity adjusted block model (PABM) and the generalized random dot product graph

@@

as in Definition 1. 
Then $A \sim GRDPG_{3, 1}(X U)$ and $A \sim PABM(\{(\lambda^{(kl)}\}_2)$ are 
equivalent.

@@

@@

\begin{equation} \label{eq:ssc}
\min_{c_i} \frac{1}{2} ||x_i - X_{-i} c_i||^2_2 + \lambda ||c_i||_1
\end{equation}

@@

Theorem 2 suggests that SSC is appropriate for community detection for the 
PABM. More precisely, Theorem 2 says that each community consists of a
$K$-dimensional subspace, and together the subspaces lie in $\mathbb{R}^{K^2}$. 
The natural approach then is to perform SSC on the ASE of $P$ or $A$.
\citet{noroozi2019estimation} instead applied SSC to $P$ and $A$ themselves.

Much of our work is built on top of work from 
\citet{rubindelanchy2017statistical}, in particular, the properties of the ASE 
of $A$ in relation to the ASE of $P$ built on $\{\lambda^{(kl)}\}_K$.

@@

If we decompose $P = Z I_{pq} Z^\top$, then
$\exists Q \in \mathbb{O}(p, q)$ such that $XU = Z Q$. Even if we start with 
the exact edge probability matrix, we cannot recover the "original" latent 
positions $XU$. Note that unlike in the case of the RDPG, 
$Q$ is not an orthogonal matrix. If $z_i$'s are the rows of $XU$, then 
$||z_i - z_j||^2 \neq ||Q z_i - Q z_j||^2$, and 
$\langle z_i, z_j \rangle \neq \langle Q z_i, Q z_j \rangle$. This prevents 
us from using the properties of $XU$ directly. In particular, if 
$Q \in \mathbb{O}(n)$, then we could use the fact that 
$\langle z_i, z_j \rangle = \langle Q z_i, Q z_j \rangle = 0$ if vertices $i$ 
and $j$ are in different communities.

@@

**Theorem 4**. Let $\hat{V}^{(n)} \in \mathbb{R}^{n \times K^2}$ be the matrix 
of $K^2$ eigenvectors of $A_n$ corresponding to the $K (K+1) / 2$ most positive 
eigenvalues and $K (K-1) / 2$ most negative eigenvalues with rows
$(\hat{v}_i^{(n)})^\top$. Let $(i, j)$ correspond to pairs belonging to 
different communities. Let $n \rho_n = \omega((\log n)^{4c}))$ as described in
\citet{rubindelanchy2017statistical}. Then for some $c > 0$,

\begin{equation} \label{eq:thm4}
  \max\limits_{i, j} ||(\hat{v}^{(n)}_i)^\top \hat{v}_j^{(n)}|| = 
    O_P\bigg( \frac{(\log n)^c}{n \sqrt{\rho_n}}\bigg)
\end{equation}

Thus, as $n \to \infty$, the partitioning of $\hat{B}_n$, the affinity matrix 
from OSE (Alg.1), results in perfect community detection with probability 1.

@@

**Proof of Theorem 2**. 
Let 
$$\Lambda^{(k)} = 
\begin{bmatrix} \lambda^{(k,1)} & \cdots & \lambda^{(k, K)} \end{bmatrix}
\in \mathbb{R}^{n_k \times K}$$

@@

The righthand side becomes

@@

The lefthand side becomes:

@@

**Proof of Theorem 4**.

**Definition 4** (Inradius) \cite{soltanolkotabi2012} 
\cite{pmlr-v28-wang13}. 
The inradius of a convex body $\mathcal{P}$, denoted by $r(\mathcal{P})$, is 
defined as the radius of the largest Euclidean ball inscribed in $\mathcal{P}$.
In addition, $r(X)$ for data matrix $X$ with rows $x_i^\top$ represents 
the inradius of the symmetric convex hull of $X$.

**Lemma 5.1**. 
Let $P_n = V_n \Lambda_n V_n^\top$ be the edge probability matrix of a PABM 
with $n$ vertices. Let $A_n \sim \Bernoulli(P_n)$ be the adjacency matrix 
drawn from $P_n$, and let $A_n = \hat{V}_n \hat{\Lambda}_n \hat{V}_n^\top$ be 
its approximate spectral decomposition using the $K (K+1) / 2$ most positive 
and $K (K-1) / 2$ most negative eigenvalues and their corresponding 
eigenvectors.Without loss of generality, order the rows of $V_n$ and 
$\hat{V}_n$ by community, and let $V_n^{(k)}$ and $\hat{V}_n^{(k)}$ be the 
$n_k \times K^2$ matrix of the rows of the $k$^th^ community in $V_n$ and 
$\hat{V}_n$ respectively. 
Let $\mu_n^{(k)} = \mu(V_n^{(k)})$ be the subspace incoherence of $V_n^{(k)}$ 
as defined by \citet{soltanolkotabi2012}. Then

\begin{equation} \label{eq:mu-zero}
\mu_n^{(k)} = 0
\end{equation}

with probability 1 $\forall n, k$.

@@

The basis of this theorem is Theorem 6 from \citet{pmlr-v28-wang13}, which 
states that the SSC detection property holds if the noise is small enough and 
the subspace inradius is greater than the subspace incoherence for each 
community $k$.

@@

Thus the conditions for the SSC detection property from Theorem 6 from
\citet{pmlr-v28-wang13} are satisfied almost surely as $n \to \infty$.

>>>>

---
title: "Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Community Detection and Parameter Estimation"
author: John Koo, Minh Tang, Michael Trosset
output:
  pdf_document:
    citation_package: natbib
    number_sections: yes
# output: html_document
fontsize: 12pt
geometry: "left=1in,right=1in,top=1in,bottom=1in"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square}
- \usepackage{verbatim}
bibliography: misc.bib
abstract: |
  In this paper, we connect two random graph models, the Popularity 
  Adjusted Block Model (PABM) and the Generalized Random Dot Product Graph 
  (GRDPG) and use properties established in this connection to aid in 
  community detection and parameter estimation. In particular, we note that the
  PABM can be represented as latent positions such that points within the same 
  community lie on a subspace, and the subspaces that represent each 
  community are orthogonal to one another. Using this property as well as the 
  asymptotic properties of Adjacency Spectral Embedding (ASE) of the GRDPG, we 
  are able to establish theoretical asymptotic results of our community 
  detection and parameter estimation methods for the PABM, including for 
  Sparse Subspace Clustering (SSC). 
---

@@

## The popularity adjusted block model and the generalized random dot product graph

@@

where each $\lambda^{(kl)}$ is a vector as in Definition 1. 
Then $A \sim GRDPG_{3, 1}(X U)$ and $A \sim PABM(\{(\lambda^{(kl)}\}_2)$ are 
equivalent.

@@

*Proof of Theorem 2*. 
Define the following matrices from $\{\lambda^{(kl)}\}_K$: 
$$\Lambda^{(k)} = 
\begin{bmatrix} \lambda^{(k,1)} & \cdots & \lambda^{(k, K)} \end{bmatrix}
\in \mathbb{R}^{n_k \times K}$$

@@

\begin{equation} \label{eq:ssc}
c_i = \arg\min_{c} \frac{1}{2} ||x_i - X_{-i} c||^2_2 + \lambda ||c||_1
\end{equation}

@@

**Definition 4** (Subspace Detection Property) \cite{pmlr-v28-wang13}. 
Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be noisy 
points sampled from $K$ subspaces. Let $C$ and $B$ be constructed from the 
solutions of LASSO problems as described in (\ref{eq:ssc}). If each column of 
$C$ has nonzero norm and $B_{ij} = 0$ $\forall$ $x_i$ and $x_j$ sampled from 
different subspaces, then $X$ obeys the subspace detection property.

**Remark**. In practice, sample $X$ often does not obey the subspace detection
property. In such cases, $B$ is treated as an affinity matrix for a graph which 
is then partitioned into $K$ subgraphs to obtain the clustering. On the other 
hand, if $X$ does obey the subspace detection property, $B$ describes a graph 
with at least $K$ disconnected subgraphs. Ideally, when the subspace detection
property holds, there are exactly $K$ subgraphs which map to each subspace, 
but it could be the case that some of the subspaces are represented by 
multiple disconnected subgraphs.

Theorem 2 suggests that SSC is appropriate for community detection for the 
PABM. More precisely, Theorem 2 says that each community consists of a
$K$-dimensional subspace, and together the subspaces lie in $\mathbb{R}^{K^2}$. 
The natural approach then is to perform SSC on the ASE of $P$ or $A$.
\citet{noroozi2019estimation} instead applied SSC to $P$ and $A$, foregoing 
embedding altogether.

Using results from \citet{soltanolkotabi2012}, it can be easily shown that 
the subspace detection property holds for $XU$, which is an ASE of $P$. More 
specifically, if points lie exactly on mutually orthogonal subspaces, then the 
subspace detection property will hold with probability 1, and this is exactly 
the case for the PABM (Theorem 2). Much of our work is then built on \citet{rubindelanchy2017statistical}, who describe the convergence behavior of 
the ASE of $A$ to the ASE of $P$, and \citet{pmlr-v28-wang13}, who show the 
necessary conditions for the subspace detection property to hold in noisy cases 
where the points lie near subspaces.

@@

If we decompose $P = Z I_{pq} Z^\top$, then
$\exists Q \in \mathbb{O}(p, q)$ such that $XU = Z Q$. Even if we start with 
the exact edge probability matrix, we cannot recover the "original" latent 
positions $XU$. Note that unlike in the case of the RDPG, $Q$ is not 
necessarily an orthogonal matrix. If $z_i$'s are the rows of $XU$, then 
$||z_i - z_j||^2 \neq ||Q z_i - Q z_j||^2$, and 
$\langle z_i, z_j \rangle \neq \langle Q z_i, Q z_j \rangle$. This prevents 
us from using the properties of $XU$ directly. In particular, if 
$Q \in \mathbb{O}(n)$, then we could use the fact that 
$\langle z_i, z_j \rangle = \langle Q z_i, Q z_j \rangle = 0$ if vertices $i$ 
and $j$ are in different communities.

@@

**Theorem 4**. 
Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij})$ be the affinity matrix from OSC (Alg. 1). Then $\forall$ pairs $(i, j)$ belonging to different communities,

\begin{equation} \label{eq:thm4}
\hat{B}_n^{(ij)} \stackrel{a.s.}{\to} 0
\end{equation}

@@

@@

The right-hand side becomes

@@

The left-hand side becomes:

@@

**Proof of Theorem 4**. 
By Theorem 3, if $i$ and $j$ belong to different communities and $v_n^{(i)}$ 
and $v_n^{(j)}$ are the corresponding eigenvectors of $P_n$, then 
$(v_n^{(i)})^\top v_n^{(j)} = 0$. Then by \citet{rubindelanchy2017statistical}, 
for eigenvectors of $A_n$, $\hat{v}_n^{(i)}$ and $\hat{v}_n^{(j)}$,

$$(\hat{v}_n^{(i)})^\top \hat{v}_n^{(j)} \stackrel{a.s.}{\to} 
(v_n^{(i)})^\top v_n^{(j)} = 0$$

since $W_n \hat{v}_n^{(i)} \stackrel{a.s.}{\to} v_n^{(i)}$ for each 
$i = 1, ..., n$ for some orthonormal $W_n$.

**Definition 5** (Inradius) \cite{soltanolkotabi2012} 
\cite{pmlr-v28-wang13}. 
The inradius of a convex body $\mathcal{P}$, denoted by $r(\mathcal{P})$, is 
defined as the radius of the largest Euclidean ball inscribed in $\mathcal{P}$.
In addition, $r(X)$ for data matrix $X$ with rows $x_i^\top$ represents 
the inradius of the symmetric convex hull of $X$.

**Lemma 5.1**. 
Let $P_n = V_n \Lambda_n V_n^\top$ be the edge probability matrix of a PABM 
with $n$ vertices. Let $A_n \sim \Bernoulli(P_n)$ be the adjacency matrix 
drawn from $P_n$, and let $A_n = \hat{V}_n \hat{\Lambda}_n \hat{V}_n^\top$ be 
its approximate spectral decomposition using the $K (K+1) / 2$ most positive 
and $K (K-1) / 2$ most negative eigenvalues and their corresponding 
eigenvectors.Without loss of generality, order the rows of $V_n$ and 
$\hat{V}_n$ by community, and let $V_n^{(k)}$ and $\hat{V}_n^{(k)}$ be the 
$n_k \times K^2$ matrix of the rows of the $k$^th^ community in $V_n$ and 
$\hat{V}_n$ respectively. 
Let $\mu_n^{(k)} = \mu(V_n^{(k)})$ be the subspace incoherence of $V_n^{(k)}$ 
as defined by \citet{soltanolkotabi2012}. Then $\mu_n^{(k)} = 0$ 
with probability 1 $\forall n, k$.

@@

The basis of this theorem is Theorem 6 from \citet{pmlr-v28-wang13}, which 
states that the subspace detection property holds if the noise is small enough 
and the subspace inradius is greater than the subspace incoherence for each 
community $k$.

@@

Thus the conditions for the subspace detection property from Theorem 6 from
\citet{pmlr-v28-wang13} are satisfied almost surely as $n \to \infty$.
